Uniform formula lower bounds for finding and counting satisfiable assignments

### 2022-05-26 article/main.tex
Discussion of speedup and slowdown

\newcommand{\Poly}{\textsf{P}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\NTIME}{\textsf{NTIME}}%
\newcommand{\coNTIME}{\textsf{coNTIME}}%
\newcommand{\SAT}{\textsf{SAT}}%
\newcommand{\ParSAT}{\ensuremath\bigoplus\textsf{SAT}}%
\newcommand{\SharpSAT}{\#\textsf{SAT}}%
\newcommand{\NAND}{\textsf{NAND}}%
\newcommand{\ND}{\textsf{NAND-DEPTH}}%
\newcommand{\NDL}[1]{\ensuremath\ND[ #1 \log n]}%

@@

Given a complexity class $\Cc$ and constants $k, (a_i)_{i\leq k}$, 
we consider \textit{quantified} complexity
class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
where $Q_i$ is one of $\forall, \exists$ or $\oplus$.
A computation in a quantified complexity class is defined as follows:
\begin{definition}[Quantified computation]
	A language $L$ is in $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$ 
	is defined by $k$ RAM machines $A_1, \ldots, A_k$ and a $\Cc$ machine $M$
	such that, given an input $x$ of length $n$:
	for every $i \le k$, $A_{i}$ takes as input $x_i$ along with $y_i$, 
	a string of $n^{a_{i} + o(1)}$ nondeterministic bits,
	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i+1}$ of length at most $n^{a_{i} + o(1)}$, 
	with the convention that $x_0 := x$.
	The output of the last stage, $x_{k+1}$, is passed to the machine $M$.

@@

	An input $x$ is in $L$ if the computation $x$ accepts starting from stage 1.
\end{definition}

@@

To prove that \SAT{} does not have formulas of depth less than $c\log n$, 
we reason by contradiction.
First, we assume that \SAT{} has such formulas, and the strong \NP-completeness of \SAT{}
allows us to deduce that the same is holds for all of $\NTIME[n]$, 
a result referred to as the ``Slowdown Lemma''.
\begin{lemma}[Slowdown Lemma]
	Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
	If \SAT{} is in $\Ff[c \log n]$\todo{define},
	then $\NTIME[n] \cup \coNTIME[n] \subseteq \Ff[c \log n]$.
\end{lemma}
\todo[inline]{talk about the issue with input size}

We also show how to speed-up (or rather, reduce the depth of) computations by adding quantifiers, 
using various ``Speedup Lemmas''. 
For example, for NAND formulas (which are the same as alternating AND/OR formulas), we have the following:
\begin{lemma}[Speedup Lemma]
	For every $c > 0, x \leq c$, we have:
	\[\textsf{NAND-depth}[c \log n] \subseteq (\exists n^{x/2}) (\forall n^1) \textsf{NAND-depth}[(c-x) \log n]\]
\end{lemma}

We then combine multiple application of these lemmas with various parameters,
with the goal of contradicting a known hierarchy theorem.
For example, we can prove that SAT does not have NAND formulas 
of depth less than $(2-\eps) \log n$ for any $\eps > 0$ with the following proof.
Assume that $\SAT\in \textsf{NAND-depth}[c \log n]$.
Then, we have
\begin{flalign*}
	NTIME[n^{2}]
		&\subseteq \NDL{2c}	& \text{Slowdown Lemma}\\
		&\subseteq (\exists n^{c})(\forall n^1)\NDL{0}& \text{Speedup Lemma with param. } x = 2c\\
		&\subseteq (\exists n^{c})\coNTIME[n]	& \\
		&\subseteq (\exists n^{c})\NDL{c}	& \text{Slowdown Lemma}\\
		&\subseteq \NTIME[n^c]	&
\end{flalign*}
For $c < 2$, this contradicts the Nondeterministic Time Hierarchy Theorem.
In practice, we use (integer) linear programming to find the best order of application
of the rules, and the best parameters to use.

We proceed similarly with properties of \ParSAT{} and \SharpSAT{}
to prove results for fomulas with XOR gates.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of thm}
Speedup Rule for Alternating formulas.

Slowdown Rule from algorithms for SAT assumptions:
\begin{lemma}
	Let $\Ff$ be a class of formulas closed under composition that can simulate projections in constant depth.
	If $\SAT{}\in \Ff[c \log n]$, then $\NTIME[n] \subseteq \Ff[c \log n]$.
\end{lemma}
\begin{proof}
	This follows from the fact that $\SAT{}$ is \NP-complete under uniform poly-log time first-order projections.
\end{proof}

Can we say the same for $\ParSAT$ and $\SharpSAT$?

>>>>

\newcommand{\Poly}{\textsf{P}}%
\newcommand{\NP}{\textsf{NP}}%
\newcommand{\NTIME}{\textsf{NTIME}}%
\newcommand{\coNTIME}{\textsf{coNTIME}}%
\newcommand{\SAT}{\textsf{SAT}}%
\newcommand{\ParSAT}{\ensuremath\bigoplus\textsf{SAT}}%
\newcommand{\SharpSAT}{\#\textsf{SAT}}%
\newcommand{\NAND}{\textsf{NAND}}%
\newcommand{\ND}{\textsf{NDepth}}%
\newcommand{\NDL}[1]{\ensuremath\ND[ #1 \log n]}%

@@

Given a complexity class $\Cc$ and constants $k, (a_i)_{i\leq k}$, 
we consider \textit{quantified} complexity
class $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$, 
where $Q_i$ is one of $\forall, \exists$ or $\oplus$.
A computation in a quantified complexity class is defined as follows:
\begin{definition}[Quantified computation]
	A language $L$ is in $(Q_1~n^{a_1})\ldots(Q_k~n^{a_k})~\Cc$ 
	is defined by $k$ RAM machines $A_1, \ldots, A_k$ and a $\Cc$ machine $M$
	such that, given an input $x$ of length $n$:
	for every $i \le k$, $A_{i}$ takes as input $x_i$ along with $y_i$, 
	a string of $n^{a_{i} + o(1)}$ nondeterministic bits,
	runs in time $n^{a_{i} + o(1)}$ and outputs a string $x_{i+1}$ of length at most $n^{a_{i} + o(1)}$, 
	with the convention that $x_1 := x$.
	The output of the last stage, $x_{k+1}$, is passed to the machine $M$.

@@

	An input $x$ is in $L$ if the computation $x$ accepts starting from stage 1.
\end{definition}
\todo[inline]{allow $O(\log n)$ quantifiers ?}

@@

To prove that \SAT{} does not have formulas of depth less than $c\log n$, 
we reason by contradiction.
First, we assume that \SAT{} has such formulas, and the strong \NP-completeness of \SAT{}
allows us to deduce that the same is holds for all of $\NTIME[n]$, 
a result referred to as the ``Slowdown rule''.
\begin{lemma}[Slowdown rule]
	Let $\Ff$ be an expressive\todo{define} class of formulas, and let $c > 1$.
	If \SAT{} is in $\Ff[c \log n]$\todo{define},
	then $\NTIME[n] \cup \coNTIME[n] \subseteq \Ff[c \log n]$.
\end{lemma}
\todo[inline]{talk about the issue with input size}

We also show how to speed-up (or rather, reduce the depth of) computations by adding quantifiers, 
using various ``Speedup rules''. 
For example, for NAND formulas (which are the same as alternating AND/OR formulas), we have the following:
\begin{lemma}[Speedup rule]
	For every $c > 0, x \leq c$, we have:
	\[\textsf{NAND-depth}[c \log n] \subseteq (\exists n^{x/2}) (\forall n^1) \textsf{NAND-depth}[(c-x) \log n]\]
\end{lemma}

We then combine multiple application of these lemmas with various parameters,
with the goal of contradicting a known hierarchy theorem.
For example, we can prove that SAT does not have NAND formulas 
of depth less than $(2-\eps) \log n$ for any $\eps > 0$ with the following proof.
Assume that $\SAT\in \textsf{NAND-depth}[c \log n]$.
Then, we have
\begin{flalign*}
	NTIME[n^{2}]
		&\subseteq \NDL{2c}	& \text{Slowdown rule}\\
		&\subseteq (\exists n^{c})(\forall n^1)\NDL{0}& \text{Speedup rule with param. } x = 2c\\
		&\subseteq (\exists n^{c})\coNTIME[n]	& \\
		&\subseteq (\exists n^{c})\NDL{c}	& \text{Slowdown rule}\\
		&\subseteq \NTIME[n^c]	&
\end{flalign*}
For $c < 2$, this contradicts the Nondeterministic Time Hierarchy Theorem.

In practice, we use (integer) linear programming to find the best order of application
of the rules, and the best parameters to use.

%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%
\section{Proof of main result}

\subsection{Speedup and Slowdowns}
We start by Slowdown rule from algorithms for SAT assumptions:
\begin{lemma}[Slowdown rule]\label{lemma:slowdown}
	If $\SAT{}\in \NDL{c}$, then for every $d > 1$, $\NTIME[n^d] \cup \coNTIME[n^d] \subseteq \NDL{c\cdot d}$.
\end{lemma}

A key idea for the proof of Lemma~\ref{lemma:slowdown} 
is the fact that any language in $\NTIME[n]$ can
be reduced to $\SAT{}$ via uniform poly-log time first-order projections, 
with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
Intuitively, a first order projection is a depth 0 circuit, 
i.e. each output bit is either a constant (0 or 1),
an input $x_i$ or its negation, $\neg x_i$.
For a more in-depth exposition of first-order projections,
see for example~\cite[End of Sec.~3]{allender1997first}.
\begin{proof}[Proof of Lemma~\ref{lemma:slowdown}]
	We first prove that $\SAT{}\in \NDL{c}$ implies that $\NTIME[n] \subseteq \NDL{c}$,
	then use a padding argument to show that $\forall d>1, \NTIME[n^d] \subseteq \NDL{cd}$,
	and the result follows since $\NDL{cd}$ is closed under complement.

	Let $L\in\NTIME[n]$. 
	As stated above, $L$ is Karp-reducible to \SAT{} via
	uniform poly-log time first-order projections, 
	\todo{ref for this? Folklore?}
	with instances of size $n$ reduced to formulas of size $n^{1+o(1)}$.
	Now, notice that the composition of a NAND formula of depth $d$ with a first order projection
	is also a NAND formula of depth $d$: it only consists in relabeling leaves of the formula.
	Poly-log time uniformity is also preserved. 
	Therefore, we can use $(c+o(1)) \log n$-depth NAND formulas for \SAT{} on $n^{1+o(1)}$
	inputs to decide $L$ by composing them with the first order projection; 
	the resulting formula has size $(c+o(1)) \log n^{1+o(1)} = (c+o(1)) \log n$,
	therefore $L\in\NDL{c}$.

	We now use a padding argument to lift this result to $\NTIME[n^d]$.
	Let $d>1$, let $L\in \NTIME[n^d]$ and let $M$ be a nondeterministic machine that decides $L$
	in time $n^{d+o(1)}$.
	Consider the padded language $L' = \{x1^{|x|^d - |x|} \mid x\in L\}$.
	Given an input $y$ of length $m$, we can decide whether $y\in L'$ in time 
	$m^{1+o(1)}$ as follows:
	\begin{itemize}
		\item if $y$ is not of the form $x1^{|x|^d - |x|}$ for some $x$, reject $y$.
		This can be done in time $O(m)$.
		\item otherwise, $y = x1^{|x|^d - |x|}$. Run $M$ on $x$, and accept $y$ 
		if and only it accepts $x$. This takes time $|x|^{d+o(1)} = m^{1+o(1)}$.
	\end{itemize}
	This implies $L'$ is in $\NTIME[n]$, which in turn implies that $L'\in\NDL{c}$.
	We can then use the formula descriptor $A$ for $L'$ 
	to describe formulas of depth $(c + o(1))\log \left(n^{d+o(1)}\right) = (cd + o(1))\log n$ for $L$.
	On input $x,b,n$, the descriptor machine $B$ for $L$  
	calls $A$ on $x, b, n^d$.
	If $A$ output a gate type or a constant, $B$ outputs the same information.
	If $A$ outputs a literal $x_i^*$, $B$ outputs $x_i^*$ if $i \le n$, and 1 otherwise.
\end{proof}

We need to be extra careful when applying the slowdown lemma to alternating classes:
the $i$-th computation stage takes as input a string of length $m = n^{a_i}$, not $n$.
Since the previous result only holds for $\NTIME[n^d], d > 1$, 
we need to take into
account the edge case where the running time of the $i$-th
stage is sublinear in the size of its input.
We have the following:
\begin{corollary}[Extended Slowdown rule]
	If $\SAT{}\in \NDL{c}$, then for every $k \in\NN$ and $d > 1$, we have:
	\[\ldots (Q~n^{a}) (\neg Q~n^b) \NDL{d}
	\subseteq \ldots (Q~n^{a}) (\neg Q~n^b) \NDL{c\cdot\max(d, a, b)}.\]
\end{corollary}
\begin{proof}
	TODO
\end{proof}

We now discuss
Speedup rule for Alternating formulas.

Intuitively, the only information needed to know that an OR gate evaluates
to 1 is that one of its children evaluates to 1.
The role of the existential quantifier is to guess for every OR gate in the first $(x/2)\log n$ levels,
which of its children evaluates to 1, and we only evaluate the corresponding subformula.
Moreover, making this guess reduces the number of OR gates that we will evaluate
(this number doubles every two levels, instead of doubling every level), 
therefore we only need to guess $n^{x/2}$ bits. 
We can then eliminate OR gates and replace them with their selected child.
We are left with a formula starting with $(x/2)\log n$ levels, 
followed by an OR/AND formula of depth $(c-x + o(1)) \log n$.
The first part can be considered as an AND gate on $n^{x/2}$ inputs (the smaller formulas).
This formula evaluate to 1 if and only if for every of these $n^{x/2}$
smaller formulas evaluates to true: we can check this using a universal quantifiers, 
this requires $O(\frac{x}{2}\log n)$ bits.
